---
title: "Easy data processing with R"
subtitle: "EECB EcoLunch"
author: "Guillermo Garcia Costoya"
date: "2024-02-16"
output: 
  html_document:
    theme: simplex
    toc: true
    toc_float: true
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Overview**


## **Loading necessary packages**

For this vignette, we will only need to load a single package: `tidyverse`. 

```{r}
library(tidyverse) 
```

As you can see from above, by loading the `tidyverse` you load a number of different packages that are useful for many kinds of data transformations. With one line of code, you load all the packages you need. 

> **TIP**: If you have not installed some of these packages, by running `install.packages("tidyverse")` you will be able to install them all at once. 


## **The `pivot` and `mutate` functions**

```{r}
# load forest data
load("data/forest_data.RData")

# view forest data
forest_data
```


```{r}
# pivot longer all columns
forest_data %>% 
  pivot_longer(cols = everything(), names_to = "forest", values_to = "height")
```

```{r}
# using mutate to add a year column and pivoting longer all columns except year
forest_data %>% 
  mutate(year = seq(1,100, by = 1)) %>% 
  pivot_longer(!year, names_to = "forest", values_to = "height")
```

```{r}
# pivotted longer dataset
forest_data_long <- forest_data %>% 
  mutate(year = seq(1,100, by = 1)) %>% 
  pivot_longer(!year, names_to = "forest", values_to = "height")

# pivot wider to go back to the original
forest_data_long %>% pivot_wider(names_from = forest, values_from = height)
```


```{r}
# plotting all raw data by forest
forest_data %>% 
  pivot_longer(cols = everything(), names_to = "forest", values_to = "height") %>%
  ggplot(aes(x = forest, y = height)) +
  geom_jitter(width = 0.1) +
  stat_summary(col = "red")
  
```


```{r}
# plotting time series
forest_data %>% 
  mutate(year = seq(1,100, by = 1)) %>% 
  pivot_longer(!year, names_to = "forest", values_to = "height") %>%
  ggplot(aes(x = year, y = height, col = forest)) +
  geom_point() +
  geom_smooth(method = "lm")
```


## **The `select`, `filter`, `group` and `summarise` functions**

```{r}
# load forest data
load("data/lake_data.RData")

# view forest data
lake_data
```

```{r}
# creating a new column by dividing N concentration by a 100 with `mutate`
lake_data %>% mutate(nitrogen_concentration = nitrogen / 100)
```

```{r}
# selecting only some columns with `select`
lake_data %>% dplyr::select(lake, date, nitrogen)
```

```{r}
# subset data for only (==) lake A using `filter`
lake_data %>% filter(lake == "A")
```

```{r}
# subset data excluding (!=) lake B using `filter`
lake_data %>% filter(lake != "A")
```

```{r}
# subset data for lakes "B" and "C" (%in%) using `filter`
lake_data %>% filter(lake %in% c("B", "C"))
```

```{r}
# grouping the data by lake and getting the average N and Cl concentrations using `summarise`
lake_data %>% group_by(lake) %>% summarise(N = mean(nitrogen), Cl = mean(chlorine))
```

```{r}
# plotting N concentrations over time. 
lake_data %>% 
  ggplot(aes(x = date, y = nitrogen/100, col = lake)) +
  geom_line()
```

```{r}
# plotting chlorine concentrations over time
lake_data %>% 
  ggplot(aes(x = date, y = chlorine/100, col = lake)) +
  geom_line() +
  geom_hline(yintercept = 0) # horizontal line indicating zero. 
```

```{r}
# plotting chlorine 
lake_data %>% 
  filter(chlorine > 0) %>%
  ggplot(aes(x = date, y = chlorine/100, col = lake)) +
  geom_line() +
  geom_hline(yintercept = 0)
```

## **The `map` function**

```{r}
# using the `map` function to run a linear model for each lake separately
lake_data %>% 
  split(lake_data$lake) %>% 
  map(\(df) lm(nitrogen/100 ~ date, data = df)) %>%
  map(summary) # map can also be used to show the summary of the model 
```

## **Dealing with dates with `lubridate` and a real example**

```{r}
# loading data
load("data/lizards.RData")

# showing data
lizards
```
```{r}
# loading the lubridate package
library(lubridate)

# re-formatting date column
lizards$date <- mdy(lizards$date)

# re-formatting time column
lizards$time <- hm(lizards$time)

lizards
```


```{r}
# generating a DOY (day of year column)
lizards$doy <- yday(lizards$date)

# generating a MOD (minute of day column)
lizards$mod <- hour(lizards$time)*60 + minute(lizards$time)
```

```{r}
# plotting density
lizards %>% 
  ggplot(aes(x = mod/60, fill = site)) +
  geom_density(alpha = 0.5) +
  ylab("Density") + xlab("Hour of the day")
```

```{r}
# get an hour column from the mod column using mutate, group by site and hour, get total observations and plot
lizards %>% 
  mutate(hour = round(mod/60)) %>% 
  group_by(site, hour) %>% 
  summarise(total_obs = n()) %>%
  ggplot(aes(x = hour,y = total_obs, fill = site)) +
  geom_col(position = "dodge2") +
  ylab("Total Observations") + xlab("Hour of the day")
```

